---
title: "Applied Stats 2 Lab 1"
author: "Lars Kutschinski"
format: pdf
editor: visual
---

```{r}
#| message: false
library(tidyverse)
```

```         
```

```{r}
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", 
                 skip = 2, col_types = "dcddd")
head(dm)
```

# 1. Plot the female to male ratios

```{r}
dm <- dm |> mutate(mf_ratio = Female/Male)
plot_data <- dm |> 
  filter(Age == 10|Age == 20|Age == 30|Age == 40) |> 
  select(Year, Age, mf_ratio)
ggplot(data = plot_data, aes(x = Year, y = mf_ratio, color = Age)) + 
  geom_line() + 
  theme_minimal()
```

# 2. Find the minimum age for each year

```{r}
dm |> group_by(Year) |> summarize(Age[which.min(Female)])
```

# 3. Find the standard deviation

```{r}
dm |> group_by(Age) |> summarize(across(c(Male,Female,Total), sd, na.rm = TRUE)) 

```

# 4. Calculate the weighted average mortality rate

```{r}
dp <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", skip = 1, col_types = "dcddd")

total_data <-left_join(dm,dp, by = c("Year","Age"))|>
  group_by(Year) |>
  drop_na() |>
  summarize(Avg_Male_Mortality = weighted.mean(Male.x, w=Male.y, na.rm = TRUE),
            Avg_Female_Mortality = weighted.mean(Female.x, w=Female.y, na.rm = TRUE))

ggplot(data = total_data, aes(x = Year)) +
  geom_line(aes(y = Avg_Male_Mortality, color = "Male")) +
  geom_line(aes(y = Avg_Female_Mortality, color = "Female")) +
  labs(title = "Population Weighted Average Mortality Rate",
       x = "Year",
       y = "Weighted Mortality Rate") +
  theme_dark()
```

We notice that the male weighted average mortality rate is higher than the female weighted average mortality rate. However both rates have significantly decreased over the years.

# 5. Linear Regression

```{r}
dm$Age <- as.numeric(dm$Age)

model_data <- dm |>
  filter(Age < 106,Year==2000) |>
  select(Female,Age)
linear_model <- lm(log(Female) ~ Age, data = model_data)
summary(linear_model)
```

The estimated coefficient for Age in our model is 0.086891. This coefficient can be interpreted as the rate of change of female mortality byage.
