---
title: "Week 10: Temporal data"
date: today
date-format: "DD/MM/YY"
format: pdf
execute: 
  warning: false
  message: false
---

# Child mortality in Sri Lanka

```{r}
library(tidyverse)
library(here)
library(rstan)
library(tidybayes)
library(janitor)

lka <- read_csv(here("data", "lka.csv"))
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se,
                  ymax = logit_ratio + se,
                  fill =  source), alpha = 0.1) +
  theme_bw()+
  labs(title = "Ratio of neonatal to other child mortality (logged), Sri Lanka", y = "log ratio")
```

# Fitting a linear model 

Let's firstly fit a linear model in time to these data. Here's the code to do this:

```{r, results = FALSE}
observed_years <- lka$year
years <- min(observed_years):max(observed_years)
nyears <- length(years)

stan_data <- list(y = lka$logit_ratio, year_i = observed_years - years[1]+1, 
                  T = nyears, years = years, N = length(observed_years), 
                  mid_year = mean(years), se = lka$se, P = 8)

mod <- stan(data = stan_data,
             file = here("stan", "lka_linear_me.stan"))
```

Extract the results:

```{r}
res <- mod %>% 
  gather_draws(mu[t]) %>% 
  median_qi() %>% 
  mutate(year = years[t])
```


Plot the results:

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res, aes(year, .value)) + 
  geom_ribbon(data = res, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  theme_bw()+
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Linear fit in black")
```

## Question 1

```{r}
res_new <- mod |>
  gather_draws(projected_mu[p]) |>
  median_qi() |>
  mutate(year = years[nyears]+p)

ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res, aes(year, .value)) + 
  geom_ribbon(data = res, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_new, aes(year, .value), col = 'lightseagreen') + 
  geom_ribbon(data = res_new, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = 'lightseagreen')+
  theme_bw()+
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Linear fit in black, projection in light green")
```
We added a generated quantities block in the stan file for the linear model that generates projections for the years 2014-2022. The added projections can be seen by the red line in the plot 


## Question 2

```{r}
lka_5 <- read.csv(here("data", "lka_estimated.csv"), skip = 5)
lka_5 <- lka_5 |> filter(Year > 1951) |> mutate(year = Year)
lka_5 <- clean_names(lka_5)

ratio_estimate <- rbind(res %>% select(.value, .lower, .upper, year), 
                        res_new %>% select(.value, .lower, .upper, year)) |>
                        mutate(ratio_est = 1/(1 + exp(-.value)),
                               ratio_lower = 1/(1 + exp(-.lower)),
                               ratio_upper = 1/(1 + exp(-.upper))
                               )
estimate <- left_join(lka_5, ratio_estimate, by = "year") |>
                mutate(neo_est = estimate * ratio_est,
                       neo_lower = lower_bound * ratio_lower,
                       neo_upper = upper_bound * ratio_upper)
estimate <- na.omit(estimate)

ggplot(estimate, aes(x = year)) +
  geom_line(data = subset(estimate,year <= 2014),aes(y = neo_est)) +
  geom_ribbon(data = subset(estimate,year <= 2014),aes(ymin = neo_lower, ymax = neo_upper), alpha = 0.2) +
  geom_line(data = subset(estimate,year > 2014),aes(y = neo_est), color = "lightseagreen") +
  geom_ribbon(data = subset(estimate,year > 2014) ,aes(ymin = neo_lower, ymax = neo_upper), fill = "lightseagreen", alpha = 0.2) +
  
  labs(title = "Neonatal Mortality Estimates and Projections in Sri Lanka",
       y = "Neonatal Mortality",
       x = "Year", subtitle = "Projection in light green") +
  theme_minimal()
```


## Question 3

```{r}
mod1 <- stan(data = stan_data,
             iter = 3000,
             file = here("stan", "FirstOrderRW.stan"))
```

```{r}
res1 <- mod1 |>
  gather_draws(mu[t]) |>
  median_qi() |>
  mutate(year = years[t])

res1_new <- mod1 |>
  gather_draws(projected_mu[p]) |>
  median_qi() |>
  mutate(year = years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res1, aes(year, .value)) + 
  geom_ribbon(data = res1, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res1_new, aes(year, .value), col = 'lightseagreen') + 
  geom_ribbon(data = res1_new, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = 'lightseagreen')+
  theme_bw()+
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Random walk(1) fit in black, projection in light green")
```


## Question 4

```{r}
mod2 <- stan(data = stan_data,
             iter = 3000,
             file = here("stan","SecondOrderRW.stan"))
```
```{r}
res2 <- mod2 |>
  gather_draws(mu[t]) |>
  median_qi() |>
  mutate(year = years[t])

res2_new <- mod2 |>
  gather_draws(projected_mu[p]) |>
  median_qi() |>
  mutate(year = years[nyears]+p)
```

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) + 
  theme_bw()+
  geom_line(data = res2, aes(year, .value)) + 
  geom_ribbon(data = res2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res2_new, aes(year, .value), col = 'lightseagreen') + 
  geom_ribbon(data = res2_new, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = 'lightseagreen')+
  theme_bw()+
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Random walk(2) fit in black, projection in light green")

```

## Question 5

```{r}
ggplot(lka, aes(year, logit_ratio)) +
  geom_point(aes( color = source)) + 
  geom_line(aes( color = source), lty = 2) + 
  geom_ribbon(aes(ymin = logit_ratio - se, 
                  ymax = logit_ratio + se, 
                  fill =  source), alpha = 0.1) +
  geom_line(data = res, aes(year, .value)) + 
  geom_ribbon(data = res, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res_new, aes(year, .value)) + 
  geom_ribbon(data = res_new, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  
  geom_line(data = res1, aes(year, .value)) + 
  geom_ribbon(data = res1, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res1_new, aes(year, .value), col = 'blue') + 
  geom_ribbon(data = res1_new, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = 'blue')+
  geom_line(data = res2, aes(year, .value)) + 
  geom_ribbon(data = res2, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2)+
  geom_line(data = res2_new, aes(year, .value), col = 'red') + 
  geom_ribbon(data = res2_new, aes(y = .value, ymin = .lower, ymax = .upper), alpha = 0.2, fill = 'red')+
  theme_bw()+
  labs(title = "Ratio of neonatal to under-five child mortality (logit), Sri Lanka",
       y = "logit ratio", subtitle = "Linear fit in black, RW(1) in blue and RW(2) in red")


```


## Question 6
The plot that we created in question 2 showed that mortality rates have been decreasing significantly over the years. The data with the logit ratio has not clearly been reflecting that however, having a many fluctuations, which resulted in what looks like an increasing trend. Ideally our model should predict a decrease in mortality rate (equivalently in the log ratio), which the second order random walk model clearly shows the most. Hence that model should be preferred.